---
title: "Assignment 2 - Data Analysis using R Programming"
subtitle: "Historical Stock Analysis of Axis Bank for Revenue Optimization"
author: "Group 5"
date: "2026-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This analysis uses 15-minute historical stock data (2025 only) for Axis Bank. Insights on price movements and high-volume trading help the bank optimize trading strategies and generate revenue through better information systems.

# 1. Load the Dataset

```{r}
data <- read.csv("AXISBANK_15m.csv", header = TRUE)

# Convert numeric columns
data$open   <- as.numeric(data$open)
data$high   <- as.numeric(data$high)
data$low    <- as.numeric(data$low)
data$close  <- as.numeric(data$close)
data$volume <- as.numeric(data$volume)

# Create Date column
data$Date <- as.Date(substr(data$timestamp, 1, 10))

names(data)
nrow(data)
```

# 2. Structure of Dataset

```{r}
str(data)
```

# 3. Variables in Dataset

```{r}
names(data)
```

# 4. Top 15 Rows

```{r}
head(data, 15)
```

# 5. User-Defined Function

```{r}
calc_15m_return <- function(open_price, close_price) {
  ((close_price - open_price) / open_price) * 100
}

calc_15m_return(data$open[1], data$close[1])
```

# 6. Filter Data (2025 High Volume)

```{r}
filtered_data <- data[
  format(data$Date, "%Y") == "2025" & 
  data$volume > mean(data$volume, na.rm = TRUE),
]

head(filtered_data, 5)
```

# 7. Dependent & Independent Variables

```{r}
dep_df <- filtered_data[, c("Date", "close")]
ind_df <- filtered_data[, c("Date", "volume", "open")]

# Add ID for reshaping
filtered_data$id <- 1:nrow(filtered_data)

temp_wide <- filtered_data[, c("id", "Date", "close", "volume")]

long_df <- reshape(
  temp_wide,
  direction = "long",
  varying = list(c("close", "volume")),
  v.names = "Value",
  timevar = "Metric",
  times = c("close", "volume"),
  idvar = "id"
)

# Merge dependent + independent
new_df <- merge(dep_df, ind_df, by = "Date")

head(new_df, 5)
```

# 8. Remove Missing Values

```{r}
new_df <- na.omit(new_df)
```

# 9. Remove Duplicates

```{r}
new_df <- new_df[!duplicated(new_df), ]
nrow(new_df)
```

# 10. Reorder Rows (Descending Closing Price)

```{r}
ordered_df <- new_df[order(-new_df$close), ]
head(ordered_df, 5)
```

# 11. Rename Columns

```{r}
colnames(new_df)[colnames(new_df) == "close"] <- "Closing_Price"
colnames(new_df)[colnames(new_df) == "volume"] <- "Trading_Volume"

head(new_df, 3)
```

# 12. Add New Variables

```{r}
new_df$Trading_Volume_Doubled <- new_df$Trading_Volume * 2
new_df$Total_Trade_Value <- new_df$Trading_Volume * new_df$Closing_Price

head(new_df, 3)
```

# 13. Create Training Set

```{r}
set.seed(123)

train_index <- sample(1:nrow(new_df), 0.7 * nrow(new_df))
train_set <- new_df[train_index, ]

nrow(train_set)
```

# 14. Summary Statistics

```{r}
summary(new_df)
```

# 15. Mean, Median, Mode, Range

```{r}
mean_price   <- mean(new_df$Closing_Price, na.rm = TRUE)
median_price <- median(new_df$Closing_Price, na.rm = TRUE)

get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}

mode_price <- get_mode(new_df$Closing_Price)

range_price <- range(new_df$Closing_Price, na.rm = TRUE)

stats_table <- data.frame(
  Statistic = c("Mean", "Median", "Mode", "Range (Min to Max)"),
  Value = c(
    round(mean_price, 2),
    round(median_price, 2),
    round(mode_price, 2),
    paste(round(range_price[1], 2), "to", round(range_price[2], 2))
  )
)

knitr::kable(stats_table, caption = "Summary Statistics of Closing Price")
```

# 16. Scatter Plot

```{r}
plot(
  new_df$Trading_Volume,
  new_df$Closing_Price,
  main = "Scatter Plot: Trading Volume vs Closing Price",
  xlab = "Trading Volume",
  ylab = "Closing Price",
  pch = 19,
  col = "steelblue"
)
```

# 17. Bar Plot (Average Closing Price by Weekday)

```{r}
filtered_data$Weekday <- weekdays(filtered_data$Date)

avg_close <- tapply(
  filtered_data$close,
  filtered_data$Weekday,
  mean,
  na.rm = TRUE
)

barplot(
  avg_close,
  main = "Bar Plot: Avg Closing Price by Weekday",
  xlab = "Weekday",
  ylab = "Average Closing Price",
  col = "steelblue"
)
```

# 18. Pearson Correlation

```{r}
correlation <- cor(
  new_df$Trading_Volume,
  new_df$Closing_Price,
  method = "pearson"
)

correlation
```

# Conclusion

The analysis identifies revenue opportunities for Axis Bank through high-volume trading patterns. High trading volume shows a relationship with closing price, which can be leveraged for better trading decisions and revenue optimization.

# GitHub Repository Link: https://github.com/YashikaDesul/Analytics-Programming-Visualization.git